---
title: "Lesson 1 — Formulate Statistical Investigative Questions"
tutorial:
  id: lesson-1-formulate
  version: 1.0.0
output:
  learnr::tutorial:
    progressive: false
    allow_skip: false
    theme: cerulean
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
tutorial_options(exercise.checker = NULL)
```

```{r setup-server, context="server"}
# Server-side context required by shiny_prerendered
NULL
```

## Big Idea

A **statistical investigative question** anticipates **variability** in data and can be answered with data you can **collect or consider**. We'll practice telling investigable questions from vague ones and make a first bar chart.

## Learning objectives

* Identify whether a question is statistical/investigable vs. vague/opinion.

* Rewrite a vague question into an investigable one that anticipates variability.

* Produce and label a basic bar chart from a tiny data set.

* Reflect on fairness and bias in question wording and sampling (ethics micro-prompt).

**Warm-up: What makes a statistical question?**

A good statistical question expects **different answers** from different people or items. For example, “Which snack is most popular in our class?” is investigable; “What’s the best snack?” is an opinion.

```{r quiz-warmup, echo=FALSE}
quiz(
  question("Which is *investigable with data*?",
    answer("What’s the best candy?", correct = FALSE),
    answer("Which candy is most popular in our class?", correct = TRUE),
    answer("Is chocolate good?", correct = FALSE)
  )
)
```

---

## Try it: Rewrite a vague question

Rewrite this vague question to make it investigable: **“Do people read a lot?”**
Type your revised question below. Then click Show a sample answer to compare.

```{r rewrite-free, echo=FALSE}
textInput("rewrite1", label = "Your investigable question:", placeholder = "e.g., What percent of 6th graders read 30+ minutes yesterday?")
```

Show a sample answer

* “What percentage of students in our class read for at least 30 minutes yesterday?”

* “What is the average number of minutes students in Grade 7 spent reading last weekend?”

Why they work: both **define a measure**, a **group**, and **anticipate variability**.


## First Plot (Bar Chart)

Now let’s use a tiny toy dataset to practice a bar chart. Run the code, then try the prompts below.

```{r first-plot, exercise=TRUE}
candy <- data.frame(
  student   = paste0("S", 1:5),
  chocolate = c("yes", "no", "yes", "yes", "no"),
  fruity    = c("yes", "yes", "no", "yes", "no")
)

# Try changing 'chocolate' to 'fruity' in aes(x = ...)
ggplot(candy, aes(x = chocolate)) +
  geom_bar() +
  labs(title = "Candy preference (toy data)", x = "Response", y = "Count")
```

---

## Ethics Micro-Prompt (2–3 sentences)

Consider the question: “Which snack is most popular in our class?”
If you only survey friends sitting near you, what bias might that introduce, and how could you reduce it? Write 2–3 sentences.

```{r ethics-free, echo=FALSE}
textAreaInput("ethics_txt", "If you only survey nearby friends, what bias might that introduce, and how could you reduce it?", rows = 4)
```

---

## Exit Ticket

1. Which prompt is **not** investigable?

```{r quiz-exit, echo=FALSE}
quiz(
  question("Pick the *non-investigable* question:",
    answer("Is pizza the best lunch?", correct = TRUE),
    answer("What percent of students ate pizza for lunch today?", correct = FALSE),
    answer("How many different lunches were brought today?", correct = FALSE)
  )
)
```

2. One sentence: how did you change a vague question to make it investigable?
```{r}
textAreaInput("exit_sentence", "Your sentence:", rows = 2, placeholder = "I defined a group, a time frame, and a measurable quantity that varies across students.")
```

